Themis-ml: A Fairness-aware Machine Learning Interface for \\ End-to-end Discrimination Discovery and Mitigation

### 2018-01-09 paper/journal_tech_human_services/main.tex
add corrections to themis-ml paper

\documentclass{interact}
\usepackage{etoolbox}
\usepackage{graphicx}
\usepackage{hanging}
\usepackage{hyperref}
\usepackage{mdframed}
\usepackage{minted}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{apacite}
\usepackage{natbib}
\usepackage[labelfont=bf]{caption}
\bibliographystyle{apacite}
\captionsetup{labelfont = bf}
\setlength{\parskip}{1em}

@@

% define code block framing style
\BeforeBeginEnvironment{minted}{
  \begin{mdframed}[skipabove=20pt,
                   skipbelow=10pt,
                   linewidth=0pt,
                   innerleftmargin=10pt,
                   innertopmargin=0pt,
                   innerbottommargin=0pt]}
\AfterEndEnvironment{minted}{\end{mdframed}}

\title{
  Themis-ml: A Fairness-aware Machine Learning Interface for \\
  End-to-end Discrimination Discovery and Mitigation
}

% \author{
%   \name{Niels Bantilan\textsuperscript{a}\thanks{Contact: Niels Bantilan. Email: niels.bantilan@gmail.com}}
%   \affil{\textsuperscript{a}Arena.io, New York, NY}
% }

% For the draft version
\author{
  \name{Anonymous\textsuperscript{a}\thanks{Contact: anonymous. Email: anonymous}}
  \affil{\textsuperscript{a}no affiliation}
}

\maketitle

\begin{abstract}
As more industries integrate machine learning into socially sensitive decision
processes like hiring, loan-approval, and parole-granting, we are at risk of
perpetuating historical and contemporary socioeconomic disparities. This is a
critical problem because on the one hand, organizations who use but do not
understand the discriminatory potential of such systems will facilitate the
widening of social disparities under the assumption that algorithms are
categorically objective. On the other hand, the responsible use of machine
learning can help us measure, understand, and mitigate the implicit historical
biases in socially sensitive data by expressing implicit decision-making mental
models in terms of explicit statistical models. In this paper we specify,
implement, and evaluate a ``fairness-aware'' machine learning interface called
themis-ml, which is intended for use by individual data scientists and
engineers, academic research teams, or larger product teams who use machine
learning in production systems.
\end{abstract}

\begin{keywords}
Fairness-aware machine learning; discrimination discovery; statistics;
social bias; application programming interface
\end{keywords}

\section{Introduction} In recent years, the transformative potential of machine
learning (ML) in many industries has propelled ML into the forefront of
mainstream media. From improving products and services to optimizing logistics
and operations, ML and artificial intelligence more broadly offer a wide range
of tools for organizations to enhance their internal and external capabilities.

@@

Researchers have developed many discrimination discovery and fairness-aware ML
methods (\citealp{kusner2017counterfactual, kamiran2012data, kamishima2012fairness,
kamiran2012decision, zemel2013learning, zafar2017fairness, dwork2012fairness,
zliobaite2015survey}), so we build on work done by others and seek to leverage
these techniques in the context of research- and product-based machine learning
applications.

Our contributions in this paper are three-fold. First, we propose an application
programming interface (API) for ``Fairness-aware Machine Learning Interfaces''
(FMLI) in the context of a simple binary classifier. Second, we introduce
themis-ml, an FMLI-compliant library, and apply it to a hypothetical
loan-granting DSS using the German Credit Dataset \cite{bache2013uci}. Finally, we
evaluate the efficacy of themis-ml as a tool for measuring potential
discrimination (PD) in both training data and ML predictions as well as
mitigating PD using fairness-aware methods. Our hope is that themis-ml serves as
a reference implementation that others might use and extend for their own
purposes.

@@

Algorithmic bias in machine learning models can occur when a trained model
generates predictions that favor one group over another in relation to some set
of attributes, e.g. education, and some target variable, e.g. ``default on
credit''. While the definition above of bias is amoral, discrimination is in
essence moral, occurring when an action is based on biases resulting in the
systematic disenfranchisement of people based on arbitrary group membership. We
define fairness as the inverse of discrimination, meaning that a ``fairness-
aware'' model is one that produces non-discriminatory predictions.

Bias can lead to either direct (intended/explicit) or indirect
(unintended/implicit) discrimination, and the predominant legal concepts used to
determine these two types are known as disparate treatment and disparate impact,
respectively \cite{barocas2016big}. As \cite{kusner2017counterfactual,
kamiran2012data} suggest, we can address disparate treatment in ML models by
simply removing all variables that are highly correlated to the protected class
of interest, in addition to the protected class itself, from the training data.
However, as \cite{kusner2017counterfactual} points out, doing so does not
necessarily mitigate discriminatory predictions and may actually introduce
unfairness into an otherwise fair system. In contrast, addressing disparate
impact is more complex because it depends on historical processes that generated
the training data, non-linear relationships between the features and protected
class, and whether we are interested in measuring individual- or group-level
discrimination \cite{dwork2012fairness}.

@@

Table 1 outlines a simple classification system in terms of the core interfaces
in scikit-learn (sklearn), which is a machine learning library in the Python
programming language \cite{scikit-learn}, and table 2 delineates some of the
use cases that research or product teams might have to justify the use of an
FMLI.

@@

Here we propose a high-level specification of themis-ml, an open source FMLI
named after the ancient Greek titaness of justice (the library can be found on
% \href{https://github.com/cosmicBboy/themis-ml}{\underline{github}}.)
\underline{github} [\emph{link removed to maintain anonymity}].)
We adopt sklearn's principles of consistency, inspection, non- proliferation of
classes, composition, and sensible defaults \cite{buitinck2013api}, and extend
them with the following FMLI-specific principles:

\begin{hangparas}{.15in}{1}
  \textbf{Model flexibility.} Focus on fairness-aware methods that are applicable
    to a variety of model types because users might have no control or full
    control over the specific model training implementation.
\end{hangparas}
\begin{hangparas}{.15in}{1}
  \textbf{Fairness as performance.} Provide estimators and scoring metrics that
    explicitly encode a notion of both model accuracy and fairness so that models
    can optimize for both.
\end{hangparas}
\begin{hangparas}{.15in}{1}
\textbf{Transparency of fairness-utility tradeoff.} Fair models often make less
  accurate predictions [8, 13], which is an important factor when assessing
  their business impact.
\end{hangparas}

@@

\subsection{Preliminaries} In the following subsections we describe specific
methods from the ML fairness literature that map onto each of the sklearn
interfaces. Note that we only provide a high level summary of each method,
citing the original sources for more implementation details. The following
descriptions make two assumptions: (i) the positive target label \(y^{+}\)
refers to a desirable outcome, e.g. ``approve loan'', and vice versa for the
negative target label \(y^{-}\), and (ii) the protected class is a binary
variable defined as \(s \in \{d, a\}\), where \(X_d\)  are members of the
disadvantaged group and \(X_a\) are members of the advantaged group.

@@

\emph{Reweighting} takes a dataset D and assigns a weight to each observation
using conditional probabilities based on \(y\) and \(s\) \cite{kamiran2012data}.
In brief, large weights are assigned to \(X_{d, y^{+}}\) and \(X_{a, y^{-}}\) ,
while small weights are assigned to \(X_{d, y^{-}}\) and \(X_{a, y^{+}}\). The
weights are then used as input to model types that support weighted sample
observations --- which actually points to the main limitation of this method,
since not all classifiers can incorporate observation weights during the
learning process.

@@

\emph{Sampling} is composed of two methods: the first involves uniformly
sampling n observations from each group, where n is the expected size of that
group assuming a uniform distribution. The second is to preferentially sample
observations using a ranker \(R\), similar to the \emph{Relabelling} method. The
procedure is to duplicate the top-ranked \(X_{d, y^{+}}\) and \(X_{a, y^{-}}\)
while removing top-ranked \(X_{d, y^{-}}\) and \(X_{a, y^{+}}\)
\cite{kamiran2012data}.

@@

\cite{kamishima2012fairness} proposes PRR as an optimization technique that
extends the standard L1/L2-norm regularization method \cite{ng2004feature,
ribeiro2016should} by adding a prejudice index term to the objective function.
This term is equivalent to normalized mutual information, which measures the
degree to which predictions \(y\) and \(s\) are dependent on each other. With
values ranging from 0 to 1, 0 means that \(y\) and \(s\) are independent, and a
value of 1 means that they are dependent. The goal of the objective function is
to find model parameters that minimize the difference between the true label
\(y\) and the predicted label \(\hat{y}\) in addition to the degree to which
\(y\) depends on s.

@@

Themis-ml draws on two methods to make model type-agnostic predictions:
\emph{Reject Option Classification} (ROC) and \emph{Discrimination Aware
Ensemble Classification} (DAEC) \cite{kamiran2012decision}. Unlike the
Transformer and Estimator methods outlined above, ROC and DAEC do not modify the
training data or the training process. Rather, they postprocess predictions in a
way that reduces potentially discriminatory (PD) predictions.

@@

The \emph{situation test score} metric is similar to \emph{consistency}, except
we consider only \(x_i \in X_d\). This method uses \emph{mean difference} to
compute a discrimination score among neighbors \(x_j \in knn(x_i)\), producing a
score between 0 and 1, where 0 indicates no discrimination, and 1 indicates
maximum discrimination \cite{zliobaite2015survey}.

@@

In this section we use the German Credit dataset \cite{bache2013uci} to evalute
themis-ml. We use \emph{mean difference} as the ``fairness'' measure and the
\emph{area under the curve} (AUC) as the ``utility'' measure. The former
represents the degree to which PD patterns in \(D\) are learned by the ML model,
and the latter represents the predictive power of a model given the available
dataset \((X, y, s) \in D\). The following analysis is by no means meant to be a
comprehensive investigation of all possible workflows that themis- ml enables.
However, does demonstrate the potential of themis-ml as a tool that facilites
fairness-aware machine learning by enabling the user to:

@@

Using the \textbf{Scorer} interface, we measure PD patterns with respect to
\emph{credit risk} and each of the protected classes defined above using the
\emph{mean difference} and \emph{normalized mean difference} metrics.

@@

To assess the extent to which (i) a model trained on these data mirrors these PD
\emph{credit risk} distributions, and (ii) fairness-aware techniques can reduce
these methods, we used \emph{mean difference} to measure model fairness and
\emph{AUC} to measure model utility. For this experiment we specify five
conditions:

@@

The lack of reduction in \emph{mean difference} between \emph{B} and \emph{RPA},
with respect to \emph{foreign worker} and LogisticRegression, illustrates the
observation made by \cite{kusner2017counterfactual} that removing protected
attributes from the training data does not necessarily prevent the algorithm
from mirroring PD patterns in the data.

However, the sizeable reduction in \emph{mean difference} between \emph{B} and
\emph{RPA}, with respect to \emph{age below 25} and LogisticRegression model,
shows that removing protected attributes can sometimes make models more fair
while also retaining predictive power.

@@

Just as the bias-variance tradeoff has become a useful diagnostic tool to guide
ML research and application \cite{fortmann2012understanding}, the fairness-
utility tradeoff can help machine learning practitioners and researchers
determine which fairness-aware methods are suitable for their particular data
context.

@@

Future technical work might also extend the FMLI specification to include
techniques like Locally Interpretable Model-Agnostic Explanations
\cite{ribeiro2016should} and develop legal frameworks for thinking about how
different stakeholders would interact with FMLIs. For example, companies that
choose not to expose the model-training components of their internal ML
pipeline could still grant some form of access to the predictions generated by
the models if there were to be a set of standards for model transparency and
accountability.

>>>>

\documentclass[man,natbib]{apa6}
\usepackage{etoolbox}
\usepackage{graphicx}
\usepackage{hanging}
\usepackage{mdframed}
\usepackage{minted}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{apacite}
\usepackage{natbib}
\usepackage{titlesec}
\usepackage{setspace}
\usepackage[labelfont=bf]{caption}
\bibliographystyle{apacite}
\captionsetup{labelfont = bf}
\setlength{\parskip}{1em}

\titlespacing*{\subsection}
{0pt}{5.5ex plus 1ex minus .2ex}{2ex plus .2ex}

% code block single spacing
\AtBeginEnvironment{minted}{\singlespacing%
    \fontsize{10}{10}\selectfont}

% define code block framing style
\BeforeBeginEnvironment{minted}{
  \begin{mdframed}[skipabove=30pt,
                   skipbelow=20pt,
                   linewidth=0pt,
                   innerleftmargin=10pt,
                   innertopmargin=0pt,
                   innerbottommargin=0pt]}
\AfterEndEnvironment{minted}{\end{mdframed}}

\title{
  Themis-ml: A Fairness-aware Machine Learning Interface for \\
  End-to-end Discrimination Discovery and Mitigation
}
\shorttitle{Themis-ml: A Fairness-aware Machine Learning Interface}

% For the draft version
\author{Niels Bantilan}
\affiliation{Arena.io}
\affiliation{New York, NY}
\email{niels.bantilan@gmail.com}

\abstract{
As more industries integrate machine learning into socially sensitive decision
processes like hiring, loan-approval, and parole-granting, we are at risk of
perpetuating historical and contemporary socioeconomic disparities. This is a
critical problem because on the one hand, organizations who use but do not
understand the discriminatory potential of such systems will facilitate the
widening of social disparities under the assumption that algorithms are
categorically objective. On the other hand, the responsible use of machine
learning can help us measure, understand, and mitigate the implicit historical
biases in socially sensitive data by expressing implicit decision-making mental
models in terms of explicit statistical models. In this paper we specify,
implement, and evaluate a ``fairness-aware'' machine learning interface called
themis-ml, which is intended for use by individual data scientists and
engineers, academic research teams, or larger product teams who use machine
learning in production systems.
}

\keywords{
  Fairness-aware machine learning, discrimination discovery, statistics,
  social bias, application programming interface}

@@

\maketitle
\section{Introduction} In recent years, the transformative potential of machine
learning (ML) in many industries has propelled ML into the forefront of
mainstream media. From improving products and services to optimizing logistics
and operations, ML and artificial intelligence more broadly offer a wide range
of tools for organizations to enhance their internal and external capabilities.

@@

Researchers have developed many discrimination discovery and fairness-aware ML
methods (\citealp{kusner2017counterfactual, kamiran2012data, kamishima2012fairness,
kamiran2012decision, zemel2013learning, zafar2017fairness,
Dwork:2012:FTA:2090236.2090255, zliobaite2015survey}), so we build on work done
by others and seek to leverage these techniques in the context of research- and
product-based machine learning applications.

Our contributions in this paper are three-fold. First, we propose an application
programming interface (API) for ``Fairness-aware Machine Learning Interfaces''
(FMLI) in the context of a simple binary classifier. Second, we introduce
themis-ml, an FMLI-compliant library, and apply it to a hypothetical
loan-granting DSS using the German Credit Dataset (\citealp{Lichman:2013}). Finally, we
evaluate the efficacy of themis-ml as a tool for measuring potential
discrimination (PD) in both training data and ML predictions as well as
mitigating PD using fairness-aware methods. Our hope is that themis-ml serves as
a reference implementation that others might use and extend for their own
purposes.

@@

Algorithmic bias in machine learning models can occur when a trained model
generates predictions that favor one group over another in relation to some set
of attributes, e.g. education, and some target variable, e.g. ``default on
credit''. While the definition above of bias is amoral, discrimination is in
essence a moral action, occurring when biases result in the systematic
disenfranchisement of people based on arbitrary group membership. We define
fairness as the inverse of discrimination, meaning that a ``fairness- aware''
model is one that produces non-discriminatory predictions.

Bias can lead to either direct (intended/explicit) or indirect
(unintended/implicit) discrimination, and the predominant legal concepts used to
determine these two types are known as disparate treatment and disparate impact,
respectively (\citealp{barocas2016big}). As \cite{kusner2017counterfactual,
kamiran2012data} suggest, we can address disparate treatment in ML models by
simply removing all variables that are highly correlated to the protected class
of interest, in addition to the protected class itself, from the training data.
However, as \cite{kusner2017counterfactual} points out, doing so does not
necessarily mitigate discriminatory predictions and may actually introduce
unfairness into an otherwise fair system. In contrast, addressing disparate
impact is more complex because it depends on historical processes that generated
the training data, non-linear relationships between the features and protected
class, and whether we are interested in measuring individual- or group-level
discrimination (\citealp{Dwork:2012:FTA:2090236.2090255}).

@@

Table 1 outlines a simple classification system in terms of the core interfaces
in scikit-learn (sklearn), which is a machine learning library in the Python
programming language (\citealp{scikit-learn}), and table 2 delineates some of the
use cases that research or product teams might have to justify the use of an
FMLI.

@@

Here we propose a high-level specification of themis-ml, an open source FMLI
named after the ancient Greek titan of justice (the library can be found on
% \href{https://github.com/cosmicBboy/themis-ml}{\underline{github}}.)
\underline{github} [\emph{link removed to maintain anonymity}]).
We adopt sklearn's principles of consistency, inspection, non- proliferation of
classes, composition, and sensible defaults \cite{buitinck2013api}, and extend
them with the following FMLI-specific principles:

\begin{hangparas}{.15in}{1}
  \textbf{Model flexibility.} Focus on fairness-aware methods that are applicable
    to a variety of model types because users have varying degrees of control
    over the specific model training implementation.
\end{hangparas}
\begin{hangparas}{.15in}{1}
  \textbf{Fairness as performance.} Provide estimators and scoring metrics that
    explicitly encode a notion of both model accuracy and fairness so that models
    can optimize for both.
\end{hangparas}
\begin{hangparas}{.15in}{1}
\textbf{Transparency of fairness-utility tradeoff.} Fair models often make less
  accurate predictions (\citealp{kamishima2012fairness, zliobaite2015survey}),
  which is an important factor when assessing their business impact.
\end{hangparas}

@@

\subsection{Preliminaries} In the following subsections we describe specific
methods from the ML fairness literature that map onto each of the sklearn
interfaces. Note that we only provide a high level summary of each method,
citing the original sources for more implementation details. Additionally, the
themis-ml source code would also provide insights into how each method works at
a lower level. The following descriptions make two assumptions: (i) the positive
target label \(y^{+}\) refers to a desirable outcome, e.g. ``approve loan'', and
vice versa for the negative target label \(y^{-}\), and (ii) the protected class
is a binary variable defined as \(s \in \{d, a\}\), where \(X_d\)  are members
of the disadvantaged group and \(X_a\) are members of the advantaged group.

@@

\emph{Reweighting} takes a dataset D and assigns a weight to each observation
using conditional probabilities based on \(y\) and \(s\) (\citealp{kamiran2012data}).
In brief, large weights are assigned to \(X_{d, y^{+}}\) and \(X_{a, y^{-}}\) ,
while small weights are assigned to \(X_{d, y^{-}}\) and \(X_{a, y^{+}}\). The
weights are then used as input to model types that support weighted sample
observations --- which actually points to the main limitation of this method,
since not all classifiers can incorporate observation weights during the
learning process.

@@

\emph{Sampling} is composed of two methods: the first involves uniformly
sampling n observations from each group, where n is the expected size of that
group assuming a uniform distribution. The second is to preferentially sample
observations using a ranker \(R\), similar to the \emph{Relabelling} method. The
procedure is to duplicate the top-ranked \(X_{d, y^{+}}\) and \(X_{a, y^{-}}\)
while removing top-ranked \(X_{d, y^{-}}\) and \(X_{a, y^{+}}\)
(\citealp{kamiran2012data}).

@@

\cite{kamishima2012fairness} proposes PRR as an optimization technique that
extends the standard L1/L2-norm regularization method (\citealp{ng2004feature,
ribeiro2016should, park2007l1}) by adding a ``prejudice index'' term to the
objective function. This term is equivalent to normalized mutual information,
which measures the degree to which predictions \(y\) and \(s\) are dependent on
each other. With values ranging from 0 to 1, 0 means that \(y\) and \(s\) are
independent, and a value of 1 means that they are dependent. The goal of the
objective function is to find model parameters that minimize the difference
between the true label \(y\) and the predicted label \(\hat{y}\) in addition to
the degree to which \(y\) depends on s.

@@

Themis-ml draws on two methods to make model type-agnostic predictions:
\emph{Reject Option Classification} (ROC) and \emph{Discrimination Aware
Ensemble Classification} (DAEC) (\citealp{kamiran2012decision}). Unlike the
Transformer and Estimator methods outlined above, ROC and DAEC do not modify the
training data or the training process. Rather, they postprocess predictions in a
way that reduces potentially discriminatory (PD) predictions.

@@

The \emph{situation test score} metric is similar to \emph{consistency}, except
we consider only \(x_i \in X_d\). This method uses \emph{mean difference} to
compute a discrimination score among neighbors \(x_j \in knn(x_i)\), producing a
score between 0 and 1, where 0 indicates no discrimination, and 1 indicates
maximum discrimination (\citealp{zliobaite2015survey}).

@@

In this section we use the German Credit dataset (\citealp{bache2013uci}) to evaluate
themis-ml. We use \emph{mean difference} as the ``fairness'' measure and the
\emph{area under the curve} (AUC) as the ``utility'' measure. The former
represents the degree to which PD patterns in \(D\) are learned by the ML model,
and the latter represents the predictive power of a model given the available
dataset \((X, y, s) \in D\). The following analysis is by no means meant to be a
comprehensive investigation of all possible workflows that themis-ml enables.
However, it does demonstrate the potential of themis-ml as a tool that facilites
fairness-aware machine learning by enabling the user to:

@@

Using the \textbf{Scorer} interface, we measure PD patterns with respect to
\emph{credit risk} and each of the protected classes defined above using
\emph{mean difference} and \emph{normalized mean difference}.

@@

To assess the extent to which (i) a model trained on these data mirrors PD
\emph{credit risk} distributions, and (ii) fairness-aware techniques can reduce
these PD patterns, we used \emph{mean difference} to measure model fairness and
\emph{AUC} to measure model utility. For this experiment we specify five
conditions:

@@

The approximately zero reduction in \emph{mean difference} between \emph{B} and
\emph{RPA}, with respect to \emph{foreign worker} and LogisticRegression,
illustrates the observation made by \cite{kusner2017counterfactual} that
removing protected attributes from the training data does not necessarily
prevent the algorithm from mirroring PD patterns in the data.

However, the sizeable reduction in \emph{mean difference} between \emph{B} and
\emph{RPA}, with respect to \emph{age below 25} and LogisticRegression model,
shows that removing protected attributes can sometimes make models fairer
while also retaining predictive power.

@@

Just as the bias-variance tradeoff has become a useful diagnostic tool to guide
ML research and application \cite{fortmann2012understanding}, the fairness-utility
tradeoff can help machine learning practitioners and researchers
determine which fairness-aware methods are suitable for their particular data
context.

@@

Future technical work might also extend the FMLI specification to include
techniques like Locally Interpretable Model-Agnostic Explanations
(\citealp{ribeiro2016should}) and develop legal frameworks for thinking about how
different stakeholders would interact with FMLIs. For example, companies that
choose not to expose the model-training components of their internal ML
pipeline could still grant some form of access to the predictions generated by
the models if there were to be a set of standards for model transparency and
accountability.
